\chapter{Romberg Integration}
\label{chap:romberg}


\section*{Introduction}
Numerical integration (or \emph{quadrature}) is the art of estimating
\[
  I = \int_a^b f(x)\,dx
\]
when an analytic antiderivative is unavailable or inconvenient.
A good mental picture is slicing bread:
each slice has a simple shape whose area you can sum, and thinner slices
give a better approximation to the whole loaf.
Romberg integration is a systematic way of cutting those slices thinner
\emph{and} blending successive approximations to cancel leading errors,
much like stacking Russian dolls so that each larger doll hides the defects
of the smaller one inside.


\section{The Composite Trapezoidal Rule Revisited}
\label{sec:trapezoid}
\begin{definition}{Composite Trapezoidal Rule}{trapezoidal-rule}
  Given $n=2^k$ sub-intervals (\(h_k = (b-a)/2^{k}\)), the composite trapezoidal rule approximates the integral as
  \begin{equation}
    T_k = h_k\left[\frac12 f(a) +\sum_{i=1}^{2^{k}-1} f(a+i h_k) + \frac12 f(b)\right]
    = \frac{b - a}{2^{k+1}} \left[f(a) + 2\sum_{i=1}^{2^{k}-1} f(a+i h_k) + f(b)\right],
    \quad
    h_k = \frac{b-a}{2^{k}}.
  \end{equation}
\end{definition}
Its error admits an asymptotic expansion
\[
  T_k \;=\; I
  + c_2 h_k^{\,2}
  + c_4 h_k^{\,4}
  + c_6 h_k^{\,6}
  + \cdots .
\]
Think of $T_k$ as a photograph that is blurry by an amount proportional
to $h_k^{\,2}$; halving~$h_k$ halves the blur squared but leaves a faint
ghost.  We would like to remove that ghost \emph{without} taking ever-smaller
pictures.


\section{Richardson Extrapolation}
Richardson extrapolation blends two pictures of different resolutions
and subtracts the shared blur:
\[
  \mathcal{A}_{k,1}
  \;=\;
  \frac{4 T_{k} - T_{k-1}}{3},
  \qquad k\ge 1.
\]
Here $T_{k}$ uses step~$h_k$ while $T_{k-1}$ uses $2h_k$.
Because the leading $h^{2}$ error cancels,
$\mathcal{A}_{k,1}$ is \(\mathcal{O}(h_k^{\,4})\).
Each application of the extrapolation ``crank'' raises the order by~\(2\).


\section{The Romberg Table}
Romberg integration applies Richardson extrapolation \emph{recursively} on a triangular table (Table~\ref{tab:romberg}).

\begin{definition}{Romberg Method}{romberg-method}
  Let $T_k$ denote the composite trapezoidal rule with $2^k$ subintervals. The Romberg tableau entries $\mathcal{A}_{k,j}$ are defined recursively as:
    \begin{align}
      \mathcal{A}_{k,0} & = T_k, \qquad\qquad\quad\;\; k = 0, 1, \dots \\[1ex]
      \mathcal{A}_{k,j} & = \frac{4^{j} \mathcal{A}_{k,j-1} - \mathcal{A}_{k-1,j-1}}{4^{j} - 1}, \qquad j = 1, \dots, k.
    \end{align}
  Here, $k$ is the row (refinement level), and $j$ is the column (extrapolation depth). The diagonal entry $\mathcal{A}_{k,k}$ achieves error $\mathcal{O}(h_k^{2(k+1)})$.
\end{definition}
\begin{corollary}{Recursive Trapezoid update}{recursive-trapezoid}
  The first column of the Romberg table ($j=0$) can be computed recursively
  \[
    A_{k,0} = \frac12 \mathcal{A}_{k-1,0} + h_k \sum_{i=1}^{2^{k-1}}f\left(a + (2i-1)h_k\right), \qquad k \ge 1.
  \]
  This allows us to easily compute the trapezoidal rule for each row without recalculating all previous values.
\end{corollary}

Each new row halves the step size (more bread slices); each new column cancels the dominant error term (removes blur). The procedure is like climbing a staircase: horizontal steps refine the mesh, vertical steps polish the estimate.

\begin{table}[h]
  \centering
  \caption{Romberg tableau for $k_{\max}=3$.}
  \label{tab:romberg}
  \begin{tabular}{c|cccc}
    $k\backslash j$ & 0                   & 1                   & 2                   & 3                   \\ \hline
    0               & $\mathcal{A}_{0,0}$                                                                   \\
    1               & $\mathcal{A}_{1,0}$ & $\mathcal{A}_{1,1}$                                             \\
    2               & $\mathcal{A}_{2,0}$ & $\mathcal{A}_{2,1}$ & $\mathcal{A}_{2,2}$                       \\
    3               & $\mathcal{A}_{3,0}$ & $\mathcal{A}_{3,1}$ & $\mathcal{A}_{3,2}$ & $\mathcal{A}_{3,3}$ \\
  \end{tabular}
\end{table}
\emph{More bread slices} (rows) and \emph{ghost removal} (columns) lead to \textbf{high-order accuracy} at modest cost.

\section{Romberg Algorithm}
The complete Romberg algorithm consists of two main components: computing the trapezoidal rule estimates and applying Richardson extrapolation.

\begin{algorithm}[H]
  \caption{Trapezoidal Rule with $2^k$ intervals}
  \label{alg:trapezoid}
  \begin{algorithmic}[1]
    \Function{Trapezoid}{$f, a, b, n$}
      \State $h \gets (b-a)/n$
      \State $sum \gets \frac{1}{2}(f(a) + f(b))$
      \For{$i = 1$ to $n-1$}
        \State $sum \gets sum + f(a + ih)$
      \EndFor
      \State \Return $h \cdot sum$
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{Refined Trapezoidal Rule (using previous computation)}
  \label{alg:refined-trapezoid}
  \begin{algorithmic}[1]
    \Function{RefinedTrapezoid}{$f, a, b, n$}
      \State $h \gets (b-a)/n$
      \State $sum \gets 0$
      \For{$i = 1$ to $n/2$}
        \State $sum \gets sum + f(a + (2i-1)h)$
      \EndFor
      \State \Return $\frac{1}{2} \cdot T_{prev} + h \cdot sum$
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
  \caption{Complete Romberg Integration}
  \label{alg:complete-romberg}
  \begin{algorithmic}[1]
    \Function{RombergIntegration}{$f, a, b, k_{max}, \varepsilon$}
      \State Initialize array $\mathcal{A} \gets \text{zeros}(k_{max}+1, k_{max}+1)$
      \State $\mathcal{A}_{[0][0]} \gets$ \Call{Trapezoid}{$f, a, b, 1$}

      \For{$k = 1$ to $k_{max}$}
        \State $\mathcal{A}_{[k][0]} \gets$ \Call{RefinedTrapezoid}{$f, a, b, 2^k$}

        \For{$j = 1$ to $k$}
          \State $\mathcal{A}_{[k][j]} \gets \dfrac{4^j \mathcal{A}_{[k][j-1]} - \mathcal{A}_{[k-1][j-1]}}{4^j - 1}$
        \EndFor

        \If{$k > 0$ and $|\mathcal{A}_{[k][k]} - \mathcal{A}_{[k-1][k-1]}| < \varepsilon$}
          \State \Return $\mathcal{A}_{[k][k]}$
        \EndIf
      \EndFor

      \State \Return $\mathcal{A}_{[k_{max}][k_{max}]}$
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\section{Adaptive Romberg}
When $f$ is highly variable on sub-intervals, naïvely halving the global
step wastes work on easy regions.
Adaptive Romberg splits $[a,b]$ recursively, applying
Algorithm~\ref{alg:romberg} on sub-intervals until
the local error estimate meets tolerance,
much like a photographer zooming in only where the picture is blurry.

\begin{algorithm}[H]
  \caption{Romberg Integration}
  \label{alg:romberg}
  \begin{algorithmic}[1]
    \State Choose $k_{\max}$ (max depth) and tolerance $\varepsilon$
    \State $\mathcal{A}_{0,0} \gets$ \Call{Trapezoid}{$f,a,b,1$}
    \For{$k = 1$ to $k_{\max}$}
    \State $\mathcal{A}_{k,0} \gets$ \Call{RefinedTrapezoid}{$f,a,b,2^{k}$}
    \For{$j = 1$ to $k$}
    \State $\mathcal{A}_{k,j} \gets \dfrac{4^{j} \mathcal{A}_{k,j-1} - \mathcal{A}_{k-1,j-1}}{4^{j}-1}$
    \EndFor
    \If{$|\mathcal{A}_{k,k} - \mathcal{A}_{k-1,k-1}| < \varepsilon$}
    \State \Return $\mathcal{A}_{k,k}$ \Comment{converged}
    \EndIf
    \EndFor
    \State \Return $\mathcal{A}_{k_{\max},k_{\max}}$
  \end{algorithmic}
\end{algorithm}

\begin{remark}{Implementation Notes}{romberg-implementation}
  \begin{itemize}
    \item \textbf{Caching:}  $T_{k}$ re-uses all $f$ evaluations
          from $T_{k-1}$; store them to avoid recomputation.
    \item \textbf{Round-off:}  For large $k$ the increment
          $\mathcal{A}_{k,0}-\mathcal{A}_{k-1,0}$ may underflow; stop refinement when
          machine precision dominates.
    \item \textbf{Vectorisation:}  Modern CPUs/GPUs can evaluate
          $f(\mathbf{x})$ on a mesh in parallel, significantly speeding
          the first few rows.
  \end{itemize}
\end{remark}

\section{Accuracy and Convergence}
Under suitable smoothness ($f\in C^{2(k+1)}[a,b]$),
\[
  \mathcal{A}_{k,k} \;=\; I + \mathcal{O}\!\bigl(h_k^{\,2(k+1)}\bigr).
\]
Thus every diagonal step increases the algebraic accuracy by two orders.
Because $h_k=2^{-k}(b-a)$, the error decays \emph{super-geometrically}; doubling $k$ squares the convergence factor.

\begin{example}{Romberg Integration Example}{romberg}
  Integrate \(I=\int_{0}^{1} \mathrm{e}^{-x^{2}}\,dx\) (the error function
  half-integral).

  \medskip

  Table~\ref{tab:example} lists $\mathcal{A}_{k,k}$ for $k=0\dots4$.
  One sees quadratic accuracy on $T_k$ (column~0) and quartic,
  sextic, etc.\ on higher columns, reaching machine precision with
  only 31~function evaluations.

  \medskip

  \begin{center}
    \begin{tabular}{c|ccccc}
      $k$ & $\mathcal{A}_{k,0}$ & $\mathcal{A}_{k,1}$ & $\mathcal{A}_{k,2}$ & $\mathcal{A}_{k,3}$ & $\mathcal{A}_{k,4}$ \\ \hline
      0   & 0.6839397206        &                     &                     &                     &                     \\
      1   & 0.7471308777        & 0.7651801875        &                     &                     &                     \\
      2   & 0.7791906969        & 0.7853981635        & 0.7858731139        &                     &                     \\
      3   & 0.7886993009        & 0.7904804897        & 0.7907357145        & 0.7907767155        &                     \\
      4   & 0.7911868057        & 0.7913186162        & 0.7913310420        & 0.7913316310        & 0.7913316515        \\
    \end{tabular}
    \label{tab:example}
  \end{center}

  Ground truth is \(0.74682413\mbox{\,…}\); the $k=4$ diagonal estimate matches to full double-precision.
\end{example}

Romberg integration combines the simplicity of the trapezoidal rule with the power of Richardson extrapolation, delivering very high-order accuracy at modest cost for smooth integrands.

Remember the three guiding images:
\begin{enumerate}
  \item \textbf{Bread slices} (\emph{mesh refinement}).
  \item \textbf{Ghost removal} (\emph{extrapolation to cancel error}).
  \item \textbf{Staircase ascent} (\emph{iterating the process}).
\end{enumerate}
When the function is smooth, a few steps up the Romberg staircase bring you astonishingly close to the
true integral.


